---
title: "LINEaR: Assumption Checks for Simple Linear Regression"
author: "Ashini Shah"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LINEaR Demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(LINEaR)
library(ggplot2)
library(bench)
```

### Introduction to LINEaR

This vignette demonstrates the LINEaR package, which provides diagnostics for LINE assumptions (Linearity, Independence, Normality, and Equal Variance) in simple linear regression. Violation of these assumptions can invalidate estimation or inference, so transparent diagnostics are crucial. In the following, we show how to simulate data, run assumption checks, and summarize results. We also verify the correctness and performance of this package by comparing results against base R functions. 

### Generate a reproducible dataset

We use the simulate_lm_data function in the LINEaR package to generate a 100 observations from a linear model with intercept = 2, slope = 0.5, and error variance = 1. We set a seed to ensure reproducibility of the random error terms. 

```{r}
sim_data <- simulate_lm_data(
  n = 100,        # number of data points
  beta0 = 2,      # intercept parameter
  beta1 = 0.5,    # slope parameter
  sigma = 1,      # standard deviation of error term
  seed = 123      # reproducible random error
  )

head(sim_data)    # see the data; simulate_lm_data() produces n (x,y) pairs.

```

### Fit a model

We fit a simple linear model to our simulated observations using `clean_lm()`. We view a clear summary of the regression using `summary()`. Because no assumption checks have been performed yet, the output shows the usual regression summary (as seen in `lm()`) and indicates assumption tests were not run. This ensures checks are only run when requested.

```{r}
fit <- clean_lm(y ~ x, data = sim_data)
summary(fit)
```

### Perform visual diagnostics for each assumption individually

To determine whether each LINE assumption holds for this data, we generate relevant diagnostic plots.  

```{r, fig.height=6, fig.width=7}
lin_result <- check_linearity(fit)
ind_result <- check_independence(fit)
norm_result <- check_normality(fit)
eqv_result <- check_equalvariance(fit)

#View plots
lin_result$plot
ind_result$plot
norm_result$plot
eqv_result$plot
```

### Perform statistical tests for each assumption individually

To be more robust, we may also run relevant statistical tests for each assumption by setting the `use_test` parameter to `TRUE`. Tests can be used in conjunction with the plots to assess validity of each assumption.

The test results determine whether the final logical flag (*_ok) is set to `TRUE` or `FALSE`. Note that the final flag does not account for visual diagnostics; users should determine final validity using both visual results and test results together. 

`check_normality()` also has a parameter allowing for the calculation of skewness and kurtosis. It defaults to `FALSE`, but can be changed using the `use_skew_kurt` parameter.

```{r}
lin_result <- check_linearity(fit, use_test = TRUE)
ind_result <- check_independence(fit, use_test = TRUE)
norm_result <- check_normality(fit, use_test = TRUE, use_skew_kurt = TRUE)
eqv_result <- check_equalvariance(fit, use_test = TRUE)

#View test results and logical flags
lin_result$test_result        # Ramsey RESET test
lin_result$assumption_ok 

ind_result$test_result        # Durbin-Watson test
ind_result$assumption_ok

norm_result$test_result       # Shapiro-Wilk test
norm_result$assumption_ok
norm_result$skew_kurt

eqv_result$test_result        # Durbin-Watson test
eqv_result$assumption_ok

```

### Combine assumption checks with `check_LINE()`

Individual check functions can assess assumptions one at a time, while the `check_LINE()` function provides a convenient way to check all of them at once. `check_LINE()` defaults to running all four statistical tests, but users can also choose which ones to include. 

```{r}
#note the syntax change: summary calls on x$object, not x alone
fit_sum <- check_LINE(fit)$object

# summarize regression + assumptions
summary(fit_sum)
```
### Example: Skipping select tests in `check_LINE()`

Users can skip certain checks by setting arguments to `FALSE`. Skipped checks are explicitly reported as "not run" in the summary. Below, tests for independence and equal variance are skipped.

```{r}
fit_new <- check_LINE(fit, independence = FALSE, equalvariance = FALSE)$object
summary(fit_new)
```
### Example: Compact Overview with `check_LINE()`

Typing `check_LINE()` will show a concise overview of assumption results without the regression summary. For both the regression summary and assumption results, use `summary()`. 

```{r}
check_LINE(fit, independence = FALSE, equalvariance = FALSE)
```

### Failed assumption checks

If an assumption check fails (statistically), the result will be `FALSE` and users will be prompted to use visual diagnostics to draw conclusions.

```{r}
# create non-linear model
set.seed(123)
dat <- data.frame(
  x = 1:50,
  y = sqrt(1:50) + rnorm(50, sd = 0.5)
)

fit_bad <- clean_lm(y ~ x, data = dat)
fit_bad <- check_LINE(fit_bad)$object
summary(fit_bad)

```
### Errored assumption checks

If an assumption check errors, the result will be `NA`, and users will be clearly notified. In this example, the data set is perfectly linear, resulting in a failure of all checks and additional warnings. Messages communicate clearly how to proceed. 
```{r}
# Very small dataset
dat <- data.frame(x = 1:3, y = c(2, 4, 6))
fit_err <- clean_lm(y ~ x, data = dat)

# Run all checks
fit_err <- check_LINE(fit_err)$object
summary(fit_err)
```

### Comparing LINEaR against base R 

We demonstrate the correctness and efficiency of LINEaR by comparing to base R. 

```{r}
# Correctness check
fit_clean <- clean_lm(y ~ x, data = sim_data)
fit_base  <- lm(y ~ x, data = sim_data)

# Compare coefficients, residuals, fitted values (expect TRUE for all)
all.equal(coef(fit_clean$model), coef(fit_base))
all.equal(fitted(fit_clean$model), fitted(fit_base))
all.equal(residuals(fit_clean$model), residuals(fit_base))

# Efficiency check
bm <- bench::mark(
  clean_lm = { clean_lm(y ~ x, data = sim_data); NULL },
  lm       = { lm(y ~ x, data = sim_data); NULL }
)
bm$expression <- as.character(bm$expression)
bm

```

### Best Practices

- Use `check_LINE()` for quick overview; run individual checks for detailed plots.
- Use `summary(fit)` for regression + assumptions; `check_LINE()` for a compact diagnostic view.
- Combine tests and plots to draw conclusions-- don't rely on test results alone.

### Closing Comments on LINEaR

LINEaR provides a clear workflow for regression diagnostics. It combines visual plots with statistical tests to check LINE assumptions (Linearity, Independence, Normality, Equal Variance). `clean_lm()` results match base R’s `lm()`, and `summary()` integrates regression results and assumption checks into a tidy summary. The package encourages transparent and reproducible analysis by displaying clear messages and prompting visual inspection when tests are skipped or undefined. 
